---
title: "STA355 Theory of Statistical Practice"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Assignments {.tabset}

## Assignment 1

Find the handout for this assignment [here](https://drive.google.com/file/d/1VgRu9nth3I5d0G5kUUbtp780FNinUDio/view?usp=sharing)

Two datasets were used: [aircon](https://drive.google.com/file/d/141trWeUTrLHSMh6zg755j_WEJOb9bkky/view?usp=sharing) and [kevlar](https://drive.google.com/file/d/1jV1VCQc7tq17TffCZML2Q6Po4tKOIf4D/view?usp=sharing) 

My answers can be found [here](https://drive.google.com/file/d/1szecHyeL2StmZCXKDcSO4rZspWQDbWMs/view?usp=sharing)

### Technical Appendix

#### Question 1(b)
```{r a11b}
x <- rgamma(200,1)
qq.expo <- function(x,plot=T) {
  x <- sort(x)  # sorts the data from smallest to largest
  n <- length(x) # length of x
  tau <- (c(1:n)-3/8)/(n+1/4)
  if (plot) {
    plot(-log(1-tau),x,xlab="Standard exponential quantiles",
         ylab="ordered data")
    abline(a=0,b=mean(x)) # draws a line with slope equal to
    # the sample mean and intercept 0
  }
  else {
    r <- list(x=x,quantiles=-log(1-tau))
    r
  }
}
qq.expo(x)


x <- rweibull(200,1)
qq.expo <- function(x,plot=T) {
  x <- sort(x)  
  n <- length(x) 
  tau <- (c(1:n)-3/8)/(n+1/4)
  if (plot) {
    plot(-log(1-tau),x,xlab="Standard exponential quantiles",
         ylab="ordered data")
    abline(a=0,b=mean(x)) 
  }
  else {
    r <- list(x=x,quantiles=-log(1-tau))
    r
  }
}
qq.expo(x)

```

#### Question 1(c)

```{r a11c}
x <- matrix(scan("C:/Users/miche/OneDrive/Desktop/UofT/STA355/Assignment/A1/aircon.txt"),ncol=5,byrow=T)
qq.expo <- function(x,plot=T) {
  x <- sort(x)  # sorts the data from smallest to largest
  n <- length(x) # length of x
  tau <- (c(1:n)-3/8)/(n+1/4)
  if (plot) {
    plot(-log(1-tau),x,xlab="Standard exponential quantiles",
         ylab="ordered data")
    abline(a=0,b=mean(x)) # draws a line with slope equal to
    # the sample mean and intercept 0
  }
  else {
    r <- list(x=x,quantiles=-log(1-tau))
    r
  }
}
qq.expo(x)
```
#### Question 2(c)

```{r a12c}
y <- scan("C:/Users/miche/OneDrive/Desktop/UofT/STA355/Assignment/A1/kevlar.txt", character(), quote = "")
x <- y[-(1:5)]
x <- as.numeric(x)
x
x <- sort(x) 
n <- length(x) 
d <- c(n:1)*c(x[1],diff(x)) 
plot(c(1:n)/n, cumsum(d)/sum(x), xlab="t", ylab="TTT") 
abline(0,1)
```

## Assignment 2

Find the handout for this assignment [here](https://drive.google.com/file/d/1afq42oaTQxSGZO_6F18Z0uSqyaC7WCYw/view?usp=sharing) 


Two datasets were used: [geyser](https://drive.google.com/file/d/1uhPaGHTSM2_V6xFVaGCKW6a33tNTppD9/view?usp=sharing) and [incomes](https://drive.google.com/file/d/17xgYAWEceof2BUGBzVE0qxqL02egGQMA/view?usp=sharing) 

My answers can be found [here](https://drive.google.com/file/d/1gAV1f9kPqRvQi9ixmyPqo4Kw_09umd99/view?usp=sharing) 

### Technical Appendix

#### Question 1(a)
```{r a21a}
geyser <- scan("C:/Users/miche/OneDrive/Desktop/UofT/STA355/Assignment/A2/geyser.txt", character(), quote = "")
geyser <- as.numeric(geyser)
qqnorm(geyser)
shapiro.test(geyser)
```

#### Question 1(b)
```{r a21b}
plot(density(geyser),ylim=c(0,0.85))
lines(density(geyser, bw=0.05), col="blue")
lines(density(geyser, bw=0.5), col="green")
points(geyser, rep(0,272), pch="x", cex=0.5)
```

#### Question 1(c)
```{r a21c}
kde.cv <- function(x,h) {
  n <- length(x)
  if (missing(h)) {
    r <- density(x)
    h <- r$bw/4 + 3.75*c(0:100)*r$bw/100
  }
  cv <- NULL
  for (j in h) {
    cvj <- 0
    for (i in 1:n) {
      z <- dnorm(x[i]-x,0,sd=j)/(n-1)
      cvj <- cvj + log(sum(z[-i]))
    }
    cv <- c(cv,cvj/n)
  }
  r <- list(bw=h,cv=cv)
  r
}
r <- kde.cv(geyser)
plot(r$bw, r$cv)
bw <- r$bw[r$cv==max(r$cv)]
bw
plot(density(geyser, bw=bw))
```


#### Question 1(d)
```{r a21d}
x1 <- geyser[geyser<3]
x2 <- geyser[geyser>3]
length(x1)
length(x2)
mean(x1)
mean(x2)
var(x1)
var(x2)
theta <- length(x1)/272
theta
```


#### Question 2(b)
```{r a22b}
income <- scan("C:/Users/miche/OneDrive/Desktop/UofT/STA355/Assignment/A2/incomes.txt", character(), quote = "")
income <- as.numeric(income)
xi <- income[income<mean(income)]
length(xi)
mps <- length(xi)/length(income)
mps


mps.f <- function(X){
  xi <- X[X< mean(X)]
  return(length(xi)/length(X))
}
mps.f(income)

jackse <- function(theta_hat, X) {
  n <- length(X) 
  theta_minus_i <- rep(0, n) 
  for (i in 1:n) {
    theta_minus_i[i] <- theta_hat(X[-i]) 
  } 
  theta_dot <- mean(theta_minus_i) 
  jackvar <- (n-1)/n * sum((theta_minus_i - theta_dot)^2) 
  return(sqrt(jackvar)) 
  }
jackse(mps.f, income)
```


#### Question 2(d)
```{r a22d}
mps.log <- function(X){
  X <- log(X)
  xi <- X[X < (mean(X)+((var(X)^2)/2))]
  return(length(xi)/length(X))
}
mps.log(income)

qqnorm(log(income))
```


## Assignment 3

Find the handout for this assignment [here](https://drive.google.com/file/d/1L8enzZNh6NndnW-8Vh2TEDzKFdLHHnPU/view?usp=sharing) 

Two datasets were used: [geyser](https://drive.google.com/file/d/1O7MaMgrTq2_0yIQghGNRmbPg76IeVxeP/view?usp=sharing) and [bees](https://drive.google.com/file/d/1Ys2xIqX35Be_aGh8ZvYvCipF4OnmLm-V/view?usp=sharing) 

My answers can be found [here](https://drive.google.com/file/d/18yqPjvy7tFE-tHqMtniW7aqAWuYqY2xX/view?usp=sharing) 

### Technical Appendix

#### Question 1(c)
```{r a31c}
pd <- function(x, kappa,lambda){
  kappa <- sort(kappa)
  d <- diff(kappa)
  n <- length(x)
  r <- sqrt(sum(cos(x))^2+sum(sin(x))^2)
  lnp <- -lambda*kappa + log(besselI(r*kappa,0)) - n*log(besselI(kappa,0))
  lnp <- lnp - max(lnp)
  p <- exp(lnp)
  m <- length(p)
  norm <- sum(d*(p[1:(m-1)]+p[2:m])/2)
  p <- p/norm
  dd <- list(kappa=kappa, posterior=p)
  dd
}

bees <- scan("C:/Users/miche/OneDrive/Desktop/UofT/STA355/Assignment/A3/bees.txt", character(), quote = "")
bees <- as.numeric(bees)
library(NISTunits)
bees <- NISTdegTOradian(bees)
pd(bees,0,1)
```

#### Question 2(a)
```{r a32a}
geyser <- scan("C:/Users/miche/OneDrive/Desktop/UofT/STA355/Assignment/A3/geyser.txt", character(), quote = "")
geyser <- as.numeric(geyser)

shorth <- function(x, tau=1/2) {
  x <- sort(x)
  n <- length(x)
  m <- ceiling(tau*n)
  x1 <- x[1:(n-m+1)]
  x2 <- x[m:n]
  j <- c(1:(n-m+1))
  len <- x2-x1
  k <- min(j[len==min(len)])
  mean(x[k:(k+m-1)])
}

shorth(geyser)
plot(density(geyser), col="blue", lwd=3)
abline(v=4.392493, lty=3, col="red", lwd=3)
points(geyser,rep(0,272),pch=20,cex=0.5)
```


#### Question 2(b)
```{r a32b}

mean50 <- NULL
mode50 <- NULL
mean100 <- NULL
mode100 <- NULL
mean500 <- NULL
mode500 <- NULL
mean1000 <- NULL
mode1000 <- NULL
mean5000 <- NULL   
mode5000 <- NULL
for (i in 1:10000) {
  x <- rnorm(50)
  mean50 <- c(mean50,mean(x))
  mode50 <- c(mode50,shorth(x))
  x <- rnorm(100)
  mean100 <- c(mean100,mean(x))
  mode100 <- c(mode100,shorth(x))
  x <- rnorm(500)
  mean500 <- c(mean500,mean(x))
  mode500 <- c(mode500,shorth(x))
  x <- rnorm(1000)
  mean1000 <- c(mean1000,mean(x))
  mode1000 <- c(mode1000,shorth(x))
  x <- rnorm(5000)
  mean5000 <- c(mean5000,mean(x))
  mode5000 <- c(mode5000,shorth(x))
}


par(mfrow=c(2,3))
boxplot(mean50,mode50)
boxplot(mean100,mode100)
boxplot(mean500,mode500)
boxplot(mean1000,mode1000)
boxplot(mean5000,mode5000)


lnn <- log(c(50,100,500,1000,5000))
lnvar <- log(c(var(mode50),var(mode100), var(mode500), var(mode1000),var(mode5000)))
plot(lnn, lnvar, pch=20)
r <- lm(lnvar~lnn)
abline(r, col='red')
summary(r)
```


## Assignment 4

Find the handout for this assignment [here](https://drive.google.com/file/d/1POznLPhUPtxhDwg7uD1q6yCYr3axe7na/view?usp=sharing) 

One dataset was used: [bees](https://drive.google.com/file/d/1rjvASu717S2B1-22S2tCOW-SibTID5Rx/view?usp=sharing)

My answers can be found [here](https://drive.google.com/file/d/17EBif4vXJ-kpmFLhnSQxaOSmPe9kbu53/view?usp=sharing)

### Technical Appendix

#### Question 1(c)
```{r a41c}
x <- c(1:20)/21
A <- NULL
for (i in 1:20){
  y <- c(rep(0,i-1),1,rep(0,20-i))
  r <- smooth.spline(x,y,df=4)
  A <- cbind(A, r$y)
}
r4 <- eigen(A, symmetric=T)
round (r4$values,3)

A <- NULL
for (i in 1:20){
  y <- c(rep(0,i-1),1,rep(0,20-i))
  r <- smooth.spline(x,y,df=6)
  A <- cbind(A, r$y)
}
r6 <- eigen(A, symmetric=T)
round (r6$values,3)

A <- NULL
for (i in 1:20){
  y <- c(rep(0,i-1),1,rep(0,20-i))
  r <- smooth.spline(x,y,df=8)
  A <- cbind(A, r$y)
}
r8 <- eigen(A, symmetric=T)
round (r8$values,3)

A <- NULL
for (i in 1:20){
  y <- c(rep(0,i-1),1,rep(0,20-i))
  r <- smooth.spline(x,y,df=9)
  A <- cbind(A, r$y)
}
r9 <- eigen(A, symmetric=T)
round (r9$values,3)

r9 <- eigen(A,symmetric=T) 
for (i in 1:20) {
  devAskNewPage(ask = T)
  plot(x,r4$vectors[,i],type="b")
}
```


#### Question 2(e)
```{r a42e}
library(CircStats)
bees <- scan("C:/Users/miche/OneDrive/Desktop/UofT/STA355/Assignment/A4/bees.txt", character(), quote = "")
bees <- as.numeric(bees)
r.test(bees, degree=TRUE)
```
