---
title: "STA457 Time Series Analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Find the handout for this assignment [here](https://drive.google.com/file/d/1wpFwdGauKYiMHjWxoTfcg-C66OOsjOtN/view?usp=sharing) 

Teammate: Guo, Dong Lai (1000979220)

## Question 1

Derive the variance of the predictor $F_t = \delta + \sum_{j=0}^{m-2} d_j X_{t-j}$ where $\delta = 0$, $i=j+1$:    
$$ \begin{align}
\sigma^{2}_F &= var(\sum_{j=0}^{m-2} d_j X_{t-j}) \\
&= \sum_{i,j=0}^{m-2} d_i d_j Cov(X_{t-i}, X_{t-j}) \\
&= \sum_{j=0}^{m-2} d^2_j Var(X_{t-j}) + \sum_{i\neq j} d_i d_j Cov(X_{t-j},X_{t-i}) \\
&= \sum_{j=0}^{m-2} d^2_j Var(X_{t-j}) + 2\sum_{0\leq j\leq i\leq{m-2}} d_i d_j Cov(X_{t-j},X_{t-i}) \\ 
&= \sigma_x^2 \sum_{j=0}^{m-2} d^2_j + 2\sum_{j=0}^{m-3}\sum_{i=j+1}^{m-2} d_i d_j \gamma_x (i-j) \\
\end{align}$$   

Any nxn real symmetric matrix A determines a quadratic form qA in n variables by the formula:  
$q_d(d_1,d_2,...d_n)= \sum_{i=1}^{N} \sum_{j=1}^{N} d_id_j A_{i,j} = \mathbf{d^T A d}$      

Let $A_{i,j} = 2\gamma_{i-j}, \mathbf{d}=(d_1,d_2,...,d_{m-2})$, then $2\sum_{j=0}^{m-3}\sum_{i=j+1}^{m-2} d_i d_j \gamma_x (i-j) =\mathbf{d^T} A \mathbf{d}$ and where $\mathbf{d_2}$ represents the square of each value within the matrix $\mathbf{d}$.     

$$\sigma^{2}_F=\sigma_x^2 \mathbf{d_2} + \mathbf{d^T} A \mathbf{d}$$

```{r q1}
# Calculate the variance of forecaster using quadratic form
#	d: vector of dj coefficients (j=0, …, m-2)
#	X: log returns

varF<-function(d,X){
  M<-length(d)-1
  acfs<- acf(X, plot=F, type="covariance", lag.max=M)$acf
  Gamma<-toeplitz(as.vector(acfs))
  d%*%Gamma%*%as.vector(d)
  }
```

## Question 2  

Derive the expectation of the predictor $F_t = \delta + \sum_{j=0}^{m-2} d_j X_{t-j}$ where $\delta = 0$ and $d_j$ being constants, $i=j+1$:    
$$\begin{align}
\mu_F &= E(\delta+\sum_{j=0}^{m-2} d_j X_{t-j}) \\
&= \delta + \sum_{j=0}^{m-2} d_j E(X_{t-j}) \\
&= \delta + \mu \sum_{j=0}^{m-2} d_j\\
&=  \mu_x \sum_{j=0}^{m-2} d_j \\
\end{align}$$   

```{r q2}
muF<-function(d,X){mean(X)*sum(d*d)}
```


## Question 3  

Derive the autocorrelation function at lag one of the predictor, where $i=J+1$ and $\gamma(k)=\sigma^2 \sum_{j=0}^{m-2} d_j d_{j+|k|}$:  

$$\rho_F(1) = corr(X_t, F_{t-1})$$

$$\begin{align} corr(X_t, F_{t-1}) &=  \frac{Cov(X_{t}, F_{t-1})}{\sigma_x \sigma_{F_{t-1}}}\\
&=  \frac{Cov(F_{t}, F_{t-1})}{\sqrt{\gamma_x (0)}\sigma_F}\\
&= \frac{Cov(X_{t}, \sum_{i} d_i X_{t-i})}{ \sqrt{\sigma_x^2 \sum_{j=0}^{m-2} d^2_j} \quad \sqrt{\sigma_x^2 \sum_{j=0}^{m-2} d^2_j + 2\sum_{j=0}^{m-3}\sum_{i=j+1}^{m-2} d_i d_j \gamma_{i-j}}}\\
&= \frac{\sum_{i} d_i Cov(X_{t}, X_{t-i})}{\sigma_x \sqrt{\sum_{j=0}^{m-2} d^2_j} \quad \sigma_x \sqrt{\sum_{j=0}^{m-2} d^2_j + 2\sum_{j=0}^{m-3}\sum_{i=j+1}^{m-2} d_i d_j \gamma_{i-j}}}\\
&= \frac{\sum_{i=0}^{m-2} d_i \gamma_x (i)}{ \sigma_x^2 \sqrt{\sum_{j=0}^{m-2} d^2_j(\sum_{j=0}^{m-2} d^2_j + 2\sum_{j=0}^{m-3}\sum_{i=j+1}^{m-2} d_i d_j \gamma_{i-j}})} \\
\end{align}$$

In quadratic form, we let $B_{i,j}=Cov(X_{t-j},X_{t-i})$ and hence    
$$\rho_F(1)= \frac{\mathbf{d}\gamma_x(i)}{\sigma_x^2 \sqrt{\mathbf{d_2}(\mathbf{d_2}+\mathbf{d^T} A \mathbf{d})}}$$   

```{r q3}
#	Calculate ACF(1) of forecaster using matrix operation and outer function in r

rhoF<-function(d,X){
    M<-length(d)-1
    acfs<- acf(X, plot=F, type ="covariance", lag.max=M+2)$acf
    temp<-d%*%matrix(acfs[abs(outer(0:M,1:(M+1), "-"))+1],M+1, M+1)%*%as.vector(d)
    temp/varF(d,X)
}

```

## Question 4   

R function to calculate the expectation of the rule return for a given double MA trading rule, and the expected length of the holding period:
```{r q4, eval=FALSE}
corXF<-function(d,X){
  Mp<-length(d)
  acfs<- acf(X, plot=F, type= "covariance", lag.max=Mp)$acf
  sum(d*acfs[-1])/sqrt(acfs[1]*varF(d,X))
}

Hold<-function(rho){pi/acos(rho)}

# m > r >=1
d<-function(m,r){c((m-r)*((0:(r-1))+1), r*(m-(r:(m-1))-1))}

#	retX: log asset return
#	m: long-term MA
#	r: short-term MA

ruleReturn<-function(retX, m, r){
  vX<-sd(retX)
  mX<-mean(retX)
  mF<-muF(d(m,r),retX)
  vF<-sqrt(varF(d(m,r),retX))
  rXF<-corXF(d(m,r),retX)
  rF<-rhoF(d(m,r),retX)
  ER<-sqrt(2/pi)*vX*rXF*exp(-mF*mF/(vF*vF))+mX*(1-2*pnorm(-mF/vF))
  H<-Hold(rF)
  list("ER"=ER, "H"=H, "rhoF"=rF, "VF"=vF, "muF"=mF,"corXF"=rXF)
}

```


## Question 5

R function to download daily, weekly S&P500 index from 10/01/2009 to 09/30/2018 from Yahoo! Finance:  
```{r q5}
library(quantmod)
sp500 <- new.env()
getSymbols("^GSPC", env = sp500, src = "yahoo", from = as.Date("2009-10-01"), to = as.Date("2018-09-30"), adjust=TRUE)
daily <- sp500$GSPC$GSPC.Adjusted
summary(daily)

weekly <- apply.weekly(daily,last)
summary(weekly)
```

## Question 6

R Code to choose the optimal m and r for daily data:
```{r q6d, eval=FALSE}
result <- numeric(0)
m <- numeric(0) 
r <- numeric(0)
for (i in c(1,5,10,20,60,120,249)){
  for (j in c(5,10,20,60,120,250)){
    if (j>i){
      result <- c(result, ruleReturn(daily, m = j, r = i)[[1]]) 
      m <- c(m,j)
      r <- c(r,i)
    }
  }
}
m_optimal <- m[which.max(result)] 
r_optimal <- r[which.max(result)]
list(optimal_m = m_optimal, optimal_r = r_optimal)
```


R Code to choose the optimal m and r for weekly data:
```{r q6w, eval=FALSE}
numeric(0)
m <- numeric(0) 
r <- numeric(0) 
for (i in 2:51){
  for (j in (i+1):52){ 
    if (j >i){
      result <- c(result, ruleReturn(weekly, m = j, r = i)[[1]]) 
      m <- c(m,j)
      r <- c(r,i)
    }
  }
}
m_optimal <- m[which.max(result)] 
r_optimal <- r[which.max(result)]
list(optimal_m = m_optimal, optimal_r = r_optimal)
```

*Conclusion*:

- The maximum of estimated return for weekly price occurs when m=250, r=249.   

- The maximum of expected return for daily price occurs when m=52,r=51.   


## Question 7

```{r q7, eval=FALSE}
in_sample_estimate <- function(data,m, r){ 
  X <- diff(log(data))
  d <- function(j){
      if (j >= 0 & j <= r-1) {(m-r)*(j+1)}
      else if (j >= r & j <= (m-1)) {r*(m-j-1)}
  }
  f <- function(t){ 
     if (t >= m){
     output <- 0
      for (j in 0:(m-1)){
       output <- output + d(j)*X[t-j]
      }
     output
   }
    else {print('t is smaller than m')}
  }
  
  #the realized return for daily data 
  re <- numeric(0)
  for (t in (m+1):length(X)){
    re <- c(re, sign(f(t-1))*X[t])
  }
  return <- sum(re)/length(m:length(X))

  #calculate holding period 
  num_change <- 0
  for (t in (m+1):length(X)){
    if (sign(f(t)) != sign(f(t-1))) {num_change <- num_change + 1}
  }
  hold <- length(m:length(X))/num_change 
  list(return = return, holding_period = hold)
}

#Return for daily data 
in_sample_estimate(daily,250,249)[[1]] 
ruleReturn(daily,250,249)[[1]]

#Holding period for daily data 
in_sample_estimate(daily,250,249)[[2]] 
ruleReturn(daily,250,249)[[2]]

#Return for weekly data 
in_sample_estimate(weekly,52,51)[[1]]
ruleReturn(weekly,52,51)[[1]]

#Holding period for weekly data 
in_sample_estimate(weekly,52,51)[[2]]
ruleReturn(weekly,52,51)[[2]]
```

*Conclusions*

For daily data:

- The estimated return is 0.00024, while the theoretical return is 0.00332. Thus, the estimated return is smaller than the theoretical return.

- The estimated holding period is 40.3, while the theoretical holding period is 15.9. Thus, the estimated period is larger than the theoretical holding period.

For weekly data:

- The estimated return is 0.00111, while the theoretical return is 0.00159. Thus, the estimated return is smaller than the theoretical return.

- The estimated holding period is 23.2, while the theoretical return is 7.7. Thus, the estimated holding period is larger than the theoretical holding period.



